Reproducing ``Fluctuation Domains''

### 2020-09-15 rescience/article.Rmd
include intermediate tex add installation to Makefile

---
title: 'Reproducing "Fluctuation Domains"'
output:
  pdf_document:
    latex_engine: xelatex
    template: rescience.latex
    includes:
      before_body: header.tex
      after_body: suffix.tex
documentclass: rescience
bibliography: "bibliography.bib"

>>>>

---
title: 'Reproducing "Fluctuation Domains"'
output:
  pdf_document:
    keep_tex: true
    latex_engine: xelatex
    template: rescience.latex
    includes:
      before_body: header.tex
      after_body: suffix.tex
documentclass: rescience
bibliography: "bibliography.bib"

### 2020-09-15 rescience/article.tex
include intermediate tex add installation to Makefile

In this analysis I seek to reproduce the numerical results of (Boettiger,
Dushoff, and Weitz 2010), which focuses on the evolutionary trajectory
of species trait under selection, as predicted by the so-called
``Canonical Equation'' of adaptive dynamics (Dieckmann and Law 1996).
Theoretical work I present in the paper predicts the emergence of a
`flutation enhancement' regime far from an evolutionary stable point,
and suggests that the evolution of traits which start in this distant
region would deviate substantially from the expected evolutionary path
predicted by the canonical equation. In particular, the distribution of
such trajectories could be bimodal, making the ``canonical'' path far
from the most likely path. The details of this distribution cannot be
solved for analytically however, so in this paper I resorted to an
replicate exact stochastic simulations of the evolutionary model using a
Gillespie algorithm (Gillespie 1977).

@@

Notably, these versions are not entirely the same. The
GitHub/Zenodo/Dryad version has the addition of a \texttt{configure.ac}
script for autoconf, intended to make the package more portable in
assisting with the linkage to the GSL library, which is missing from the
original deposit (on the GeorgiaTech website and Dryad). The GitHub
version also lacked a copy of the \texttt{inst/} subdirectory, which
contained a copy of the manuscript \texttt{.tex} file, original
manuscript \texttt{.pdf} figures, and the R scripts that were supposed
to generate these figures. (Though both versions include a \texttt{demo}
directory that included similar code required to generate the figures,
as discussed in the next section.)

@@

The second challenge was in figuring out just what commands had to be
run to duplicate the figures. As mentioned above, the GitHub version of
the source-code did not have the manuscript source, and even the version
that did have the manuscript source did not write the manuscript itself
as a dynamic document (e.g.~in Sweave (Xie 2015)). In that version,
files \texttt{fig1.R} and \texttt{fig2.R} are were included in the same
subdirectory as the \texttt{pdf} figures referenced by the \texttt{tex}
version of the manuscript, but these files
contained only the code required to actually generate the figures
themselves once the result data had been pre-computed and saved as
\texttt{.txt} files. The result data files had also not been archived
separately. Fortunately, re-reading my own documentation, I was able to
deduce that these text files were generated by running the function
\texttt{logistic()} with the parameter settings given in the caption of
Fig 2 in the manuscript.

@@

\KeywordTok{landscape}\NormalTok{(logistic_curve, }\DecValTok{-3}\NormalTok{, }\DecValTok{3}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/fig1-1.pdf}
\caption{The three panels of figure 1, showing examples of evolutionary
landscapes and the corresponding fluctuation domains for three classic
models (logistic, chemostat, branching model). This figure is largely
theoretical, and so easily replicates Figure 1 of the original paper.
The first panel showing the logistic model depicts the fluctuation
regimes for the model parameters used in the numerical analysis.}
\end{figure}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{landscape}\NormalTok{(chemostat_curve, }\DecValTok{1}\NormalTok{, }\DecValTok{9}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/fig1-2.pdf}
\caption{The three panels of figure 1, showing examples of evolutionary
landscapes and the corresponding fluctuation domains for three classic
models (logistic, chemostat, branching model). This figure is largely
theoretical, and so easily replicates Figure 1 of the original paper.
The first panel showing the logistic model depicts the fluctuation
regimes for the model parameters used in the numerical analysis.}
\end{figure}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{landscape}\NormalTok{(branching_curve, }\FloatTok{-.8}\NormalTok{, }\FloatTok{.8}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/fig1-3.pdf}
\caption{The three panels of figure 1, showing examples of evolutionary
landscapes and the corresponding fluctuation domains for three classic
models (logistic, chemostat, branching model). This figure is largely
theoretical, and so easily replicates Figure 1 of the original paper.
The first panel showing the logistic model depicts the fluctuation
regimes for the model parameters used in the numerical analysis.}
\end{figure}

Figure 2 of the original paper has the core computational results
demonstrating the consequences of these fluctuation regimes. The code
for this analysis was implemented as pure C code, which can be run in
stand-alone mode and generates three output text files. The R package
uses the \texttt{.C} interface to pass parameters to the C function
(\texttt{Rlogistic}) which runs the ensemble and writes its output as
text. The C function does not expose a random seed, so re-running the
analysis creates results which are not bitwise identical each time. Of
course the scientific results are expected to be independent of the seed
anyway, so it is perhaps reassuring that as long as the ensemble of
replicates is large enough, the same pattern can be reproduced quite
closely. The original manuscript uses ensembles of 100,000 replicates,
which I have reproduced here. A possible source of confusion is that the
code defaults to only 100 replicates if the ensemble size is not
explicitly verified -- which is an ensemble small enough to still see
random deviations from the predicted results. I probably set that lower
default to facilitate faster testing. Likewise, the MAXTIME parameter
has to be extended from the default setting to allow the third initial
condition set to run sufficiently long.

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-2-1.pdf}
\caption{A starting condition of Xo = 1, at the edge of the fluctuation
dissipation regime. The whole evolutionary trajectory thus occurs within
the dissipation regime, and mean dynamics closely match that of the
canonical equation, and show a tight normal distribution around the
expected path. This matches the trajectories seen figure 2 of the
original manuscript.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-3-1.pdf}
\caption{A starting condition of Xo = 2, matching the trajectories seen
figure 2 of the original manuscript. Deviations from the canonical path
are visible, but the distribution is not yet bimodal.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{Starting deep in the predicted fluctuation enhancement regime,
Xo = 3, the resulting distribution of trajectories is bimodal, just as
shown in the original figure 2 of the original manuscript.}
\end{figure}

@@

\begin{itemize}
\tightlist
\item
  The C methods are expected to bind R's random number generator
  engines, and not use their own methods, which allows them to
  coordinate with standard R methods for setting seed and so forth.\\
\item
  Data should be passed by reference back to R and not serialized to
  disk (though this can introduce memory constraints).\\
\item
  It is now also standard practice to register the C routines
  explicitly.\\
\item
  \texttt{SystemRequirements} field was not provided in the package
  \texttt{DESCRIPTION}. By adding this to list \texttt{libgsl}, it is
  possible to have the package successfully install on the \texttt{rhub}
  (Csárdi and Salmon 2019) testing platforms.
\end{itemize}

Note that these tighter integrations would also prevent the C libraries
from running as purely stand-alone implementation. In any event, meeting
all the automated checks of a formal package is not an expectation of
the compendium approach, which merely borrows some of the package
tooling.

@@

In conclusion, I have been able to successfully replicate the original
results. As the original results were not performed with a fixed random
seed, replicated results are statistically comparable but not bitwise
identical. In a scientific sense this is perhaps more useful than
replication with a fixed seed. My compendium approach did not install
entirely out of the box, as an autoconf configure script first required
updating. Ironically, the original version of the code published on a
co-author's personal website and linked from the manuscript both still
existed and did install out-of-the-box. Reproducing the results of the
paper required some careful reading of parameters listed in the paper
and manually knowing what functions to call, as this paper did not
include a complete ``dynamic document'' (Xie 2014, 2015; Xie, Allaire,
and Grolemund 2018) approach discussed in Marwick, Boettiger, and Mullen
(2018) and also used in this replication paper. I think this experience
underscored the critical role dependency management has in frustrating
reproducibility: this replication was relatively simple in part due to
the low number of dependencies and the technical simplicity of the code.
The other take-away for me is how easy it is to leave out small but
important steps needed to reproduce a figure when not using the dynamic
document approach. Despite having R scripts called ``figure1.R'' and
``figure2.R'', these scripts could not simply be run by themselves to
create the figures.

@@

\leavevmode\hypertarget{ref-rmarkdown}{}%
Xie, Yihui, J.J. Allaire, and Garrett Grolemund. 2018. \emph{R Markdown:
The Definitive Guide}. Boca Raton, Florida: Chapman; Hall/CRC.
\url{https://bookdown.org/yihui/rmarkdown}.

>>>>

In this analysis I seek to reproduce the numerical results of
(Boettiger, Dushoff, and Weitz 2010), which focuses on the evolutionary
trajectory of species trait under selection, as predicted by the
so-called ``Canonical Equation'' of adaptive dynamics (Dieckmann and Law
1996). Theoretical work I present in the paper predicts the emergence of
a `fluctuation enhancement' regime far from an evolutionary stable
point, and suggests that the evolution of traits which start in this
distant region would deviate substantially from the expected
evolutionary path predicted by the canonical equation. In particular,
the distribution of such trajectories could be bimodal, making the
``canonical'' path far from the most likely path. The details of this
distribution cannot be solved for analytically however, so in this paper
I resorted to an replicate exact stochastic simulations of the
evolutionary model using a Gillespie algorithm (Gillespie 1977).

@@

Notably, these versions are not entirely the same. The
GitHub/Zenodo/Dryad version has the addition of a \texttt{configure.ac}
script for autoconf, intended to make the package more portable in
assisting with the linkage to the GSL library, which is missing from the
original deposit (on the GeorgiaTech website and Dryad). The GitHub
version also lacked a copy of the \texttt{inst/} sub-directory, which
contained a copy of the manuscript \texttt{.tex} file, original
manuscript \texttt{.pdf} figures, and the R scripts that were supposed
to generate these figures. (Though both versions include a \texttt{demo}
directory that included similar code required to generate the figures,
as discussed in the next section.)

@@

The second challenge was in figuring out just what commands had to be
run to duplicate the figures. As mentioned above, the GitHub version of
the source-code did not have the manuscript source, and even the version
that did have the manuscript source did not write the manuscript itself
as a dynamic document (e.g.~in Sweave (Xie 2015)). In that version,
files \texttt{fig1.R} and \texttt{fig2.R} are were included in the same
sub-directory as the \texttt{pdf} figures referenced by the \texttt{tex}
version of the manuscript, but these files contained only the code
required to actually generate the figures themselves once the result
data had been pre-computed and saved as \texttt{.txt} files. The result
data files had also not been archived separately. Fortunately,
re-reading my own documentation, I was able to deduce that these text
files were generated by running the function \texttt{logistic()} with
the parameter settings given in the caption of Fig 2 in the manuscript.

@@

\KeywordTok{landscape}\NormalTok{(logistic_curve, }\DecValTok{-3}\NormalTok{, }\DecValTok{3}\NormalTok{)}
\KeywordTok{landscape}\NormalTok{(chemostat_curve, }\DecValTok{1}\NormalTok{, }\DecValTok{9}\NormalTok{)}
\KeywordTok{landscape}\NormalTok{(branching_curve, }\FloatTok{-.8}\NormalTok{, }\FloatTok{.8}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\includegraphics{article_files/figure-latex/fig1-1} \includegraphics{article_files/figure-latex/fig1-2} \includegraphics{article_files/figure-latex/fig1-3} \caption{The three panels of figure 1, showing examples of evolutionary landscapes and the corresponding fluctuation domains for three classic models (logistic, chemostat, branching model). This figure is largely theoretical, and so easily replicates Figure 1 of the original paper.  The first panel showing the logistic model depicts the fluctuation regimes for the model parameters used in the numerical analysis.}\label{fig:fig1}
\end{figure}

Figure 2 of the original paper has the core computational results
demonstrating the consequences of these fluctuation regimes. The code
for this analysis was implemented as pure C code, which can be run in
stand-alone mode and generates three output text files. The R package
uses the \texttt{.C} interface to pass parameters to the C function
(\texttt{Rlogistic}) which runs the ensemble and writes its output as
text. The C function does not expose a random seed, so re-running the
analysis creates results which are not bit-wise identical each time. Of
course the scientific results are expected to be independent of the seed
anyway, so it is perhaps reassuring that as long as the ensemble of
replicates is large enough, the same pattern can be reproduced quite
closely. The original manuscript uses ensembles of 100,000 replicates,
which I have reproduced here. A possible source of confusion is that the
code defaults to only 100 replicates if the ensemble size is not
explicitly verified -- which is an ensemble small enough to still see
random deviations from the predicted results. I probably set that lower
default to facilitate faster testing. Likewise, the MAXTIME parameter
has to be extended from the default setting to allow the third initial
condition set to run sufficiently long.

@@

\begin{figure}
\centering
\includegraphics{article_files/figure-latex/unnamed-chunk-2-1.pdf}
\caption{A starting condition of Xo = 1, at the edge of the fluctuation
dissipation regime. The whole evolutionary trajectory thus occurs within
the dissipation regime, and mean dynamics closely match that of the
canonical equation, and show a tight normal distribution around the
expected path. This matches the trajectories seen figure 2 of the
original manuscript.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{article_files/figure-latex/unnamed-chunk-3-1.pdf}
\caption{A starting condition of Xo = 2, matching the trajectories seen
figure 2 of the original manuscript. Deviations from the canonical path
are visible, but the distribution is not yet bimodal.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{article_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{Starting deep in the predicted fluctuation enhancement regime,
Xo = 3, the resulting distribution of trajectories is bimodal, just as
shown in the original figure 2 of the original manuscript.}
\end{figure}

@@

\begin{itemize}
\tightlist
\item
  The C methods are expected to bind R's random number generator
  engines, and not use their own methods, which allows them to
  coordinate with standard R methods for setting seed and so forth.
\item
  Data should be passed by reference back to R and not serialized to
  disk (though this can introduce memory constraints).
\item
  It is now also standard practice to register the C routines
  explicitly.
\item
  \texttt{SystemRequirements} field was not provided in the package
  \texttt{DESCRIPTION}. By adding this to list \texttt{libgsl}, it is
  possible to have the package successfully install on the \texttt{rhub}
  (Csárdi and Salmon 2019) testing platforms.
\end{itemize}

Note that these tighter integration would also prevent the C libraries
from running as purely stand-alone implementation. In any event, meeting
all the automated checks of a formal package is not an expectation of
the compendium approach, which merely borrows some of the package
tooling.

@@

In conclusion, I have been able to successfully replicate the original
results. As the original results were not performed with a fixed random
seed, replicated results are statistically comparable but not bit-wise
identical. In a scientific sense this is perhaps more useful than
replication with a fixed seed. My compendium approach did not install
entirely out of the box, as an autoconf configure script first required
updating. Ironically, the original version of the code published on a
co-author's personal website and linked from the manuscript both still
existed and did install out-of-the-box. Reproducing the results of the
paper required some careful reading of parameters listed in the paper
and manually knowing what functions to call, as this paper did not
include a complete ``dynamic document'' (Xie 2014, 2015; Xie, Allaire,
and Grolemund 2018) approach discussed in Marwick, Boettiger, and Mullen
(2018) and also used in this replication paper. I think this experience
underscored the critical role dependency management has in frustrating
reproducibility: this replication was relatively simple in part due to
the low number of dependencies and the technical simplicity of the code.
The other take-away for me is how easy it is to leave out small but
important steps needed to reproduce a figure when not using the dynamic
document approach. Despite having R scripts called ``figure1.R'' and
``figure2.R'', these scripts could not simply be run by themselves to
create the figures.

@@

\leavevmode\hypertarget{ref-rmarkdown}{}%
Xie, Yihui, J. J. Allaire, and Garrett Grolemund. 2018. \emph{R
Markdown: The Definitive Guide}. Boca Raton, Florida: Chapman; Hall/CRC.
\url{https://bookdown.org/yihui/rmarkdown}.
